Data Management \\ and LSST Special Programs

### 2017-11-14 DMTN-065.tex
Add DMTN handle to tex file
Use more modern latex

\setDocRef{DMTN-nnn}
\date{\today}
\setDocRevision{TBD}
\setDocStatus{draft}

\setDocAbstract{{\bf This is a draft document and may yet change.} The purpose of this study is to ensure that the LSST Data Management's plans for processing data from Special Programs (Deep Drilling Fields and Mini-Surveys) will meet the needs of the scientific community, and that these plans have been appropriately sized for computational processing and are accurately represented in the work-hours budget. Unresolved issues spawn {\bf Actions}, which are typically documentation change requests or proposed studies to assess a potential problem.}

@@

\noindent The purpose of this study is to ensure that the requirements and plans of DM are accurately expressed in all major DM documentation, and that the computational system and human work-hours needed to accomplish Special Programs processing have been accurately evaluated. This study spawns {\bf Action} items, which are typically either documentation change requests or proposals for future studies needed to resolve an issue.

%In particular, this study is designed to review: (1) the processing and products that are required to enable science for Special Programs, (2) the current plans for DM processing and products with respect to Special Programs, (3) whether the DM plans meet the expected needs of Special Programs, and (4) any necessary changes to be written into the requirements, designs, and plans, in which case an {\bf Action} is spawned.

@@

Section \ref{sec:docrev} steps through each of the relevant LSST documents and highlights specific requirements or text in these documents that might need additions or clarifications to properly reflect DM's plans for processing Special Programs data.
%These {\bf Concerns} are not prioritized by urgency or importance, but instead represent more of a complete set of possible issues. {\bf The Concerns raised by this study fall into two main categories:} \\
%1. Whether DM will be able to process the full diversity of non-standard visit images, and if not, what the boundaries will be. \\
%2. Ensuring that DM's intent to process Special Programs data with reconfigured pipelines, and to provide separate data products that meet the science goals of each program, is accurately reflected in all documents, as it is in the DPDD, and is in the staffing budget. \\

@@

All aspects related to DM's {\it processing} of Special Programs data are addressed in Section \ref{sec:dmplans}.

@@

{\bf Shorter Exposures.} The camera constraint on the minimum supported exposure time is currently 1 second (stretch goal 0.1 seconds). The minimum exposure time for an image to be successfully reduced with Instrument Signature Removal (ISR) will be a DM constraint. Assuming that 1 second exposure can be reduced and calibrated, its detected point sources will span a dynamic range of $r \approx 12.9$ -- $21.0$ magnitudes. A template image built on $15$ second exposures will saturate at $r \approx 15.8$, but this still leaves stars between $15.8$--$21.0$ magnitudes to be used in the PSF-matching (and all other filters have a similarly large overlap). However, in order for an image to be successfully PSF-matched to the template, the PSF must be well formed (i.e., no speckle pattern), and have a spatial variance that the pipeline is capable of modeling (i.e., be smoothly varying on some minimal scale). As a simple demonstration, MLG has made Figure \ref{fig:expt} to show that perhaps exposure times shorter than 2 seconds do not have a well-formed PSF (using the centroid of a 2D Gaussian fit as a proxy for "well-formed"). Another concern is that the Level 1 pipeline would not be able to deliver the required $60$ second latency if the camera produces more than 1 image in 30 seconds, which would disqualify a contribution to the Alert Stream.

@@

{\bf Longer Exposures.} There is no maximum exposure time specified for an LSST image. Given that the template image will be a stack of at least a year or two of data, processing a $5$--$10$ times deeper single image through the difference imaging pipeline should be fine. However, a $2\times150$ second exposure would saturate at $r \approx 18.3$, and cosmic-ray rejection completeness might suffer (unknown), which could impact the quality of a difference image and the detected sources. With a longer exposure time the $60$ second latency requirement could still be met, but any system qualities that vary on short (but $>30$ second) timescales could inhibit photometric calibration (e.g., tracking).

In conversation with DM-AP team members (Reiss, Findeisen, Connolly, Bo) there has not yet been a study of the safe range of exposure times that will be allowed to contribute to the Level 1 and Alert Stream. One possibly useful study is Chang et al. (2012), "Atmospheric point spread function interpolation for weak lensing in short exposure imaging data". They show that a 15 second exposure contains PSF variability on short spatial scales across a 1 square degree image which, for extragalactic fields with few stars (i.e., but good for weak lensing), is hard to characterize. They present a new software package to do mitigate the effects. Alternatively, we may need to use software packages {\tt PhoSim} (Peterson et al. 2015; \url{http://adsabs.harvard.edu/abs/2015ApJS..218...14P}) or {\tt ARROYO} (\url{http://proceedings.spiedigitallibrary.org/proceeding.aspx?articleid=848256}) to at least simply characterize the PSF stability as a function of exposure time.

@@

OSS-REQ-0291 requires that short exposures be spaced by 15 seconds, and so DM's ingest rate of exposures should never be more than 1 per 15 seconds, or a visit rate of 1 per 30 seconds. However, {\it if a higher image acquisition rate is cleared by the camera team}, then DM might encounter some issues such as the minimum time required to save an image to disk, the handling of a processing backup, and/or inconsistently associated {\tt DIASources}.

{\bf Minimum Save Time:} If short exposures are taken in an unbuffered sequence, the minimum time between exposures is the shutter-crossing time (1 second) plus the readout time (2 seconds): 3 seconds. This is too rapid an acquisition rate for processing with Level 1, but what if the science goals accept saving them to disk and analyzing them later? {\bf How long does it take to save an image to disk?}

{\bf Processing backup:} Are there options for short-term increases in parallel processing power at NCSA? (Such options might be needed anyway to process crowded fields with $>$10k {\tt DIASources}.) The bandwidth needed to load templates at a faster rate is also a concern, since templates have twice as many pixels, if the data acquistion rate is surveying at twice the WFD main survey rate, that's a $4\times$ additional bandwidth load. However, at NCSA there should be ways to elastically change the amount of processing power available. This is also an issue for crowded fields (below).

{\bf Unassociated {\tt DIASources}:} the consistency of {\tt DIAObjects} suffers if the same patch of sky is observed in quick succession (overlapping regions of short exposures and/or sustained 30 second visits to the same field). In other words, it is not guaranteed that the next string of processing would have access to the latest {\tt DIAObjects} for association. It is furthermore unclear whether we can flag an Alert as potentially having incomplete source-object associations. This is also an issue for crowded fields (below).

@@

K.-T. Lim has pointed out that an odd number of exposures is a non-standard visit; two snaps is hardwired into the code. This is baked-in to a configuration so that the pipeline can have a definition of what kind of timing delay constitutes ``late".  Moving away from 2 exposures per visit requires a configuration change to the pipelines, which incurs an overhead (up to 1 minute) -- in fact, K.-T. things that between $10$ and $120$ seconds exposure times can easily be handled by the pipeline (i.e., can be run through ISR using scaled calibration frames), so long as they come in pairs. The real problem is knowing how long the processing should take, and not killing a process that is taking longer because there were 4 snaps in the visit instead of 2. To accommodate non-standard visits requires that the scheduler pass on the information of the number of snaps in the visit (\ref{DMSR-1}). Then the processing pipeline will know to, e.g., not attempt to difference the two snaps in the case were there is an odd number of snaps in a visit. {\it MLG -- I've heard rumors of a CR regarding alternate standard visits of $1\times30$ seconds, but do not know the status or implications of this.}

@@

Special Programs are more likely to include crowded fields than the WFD main survey area. Due to the increased number of sources, the number of {\tt DIASources} -- and therefore the number of Alerts -- increases as well. In turn, this increases the processing time and in some cases, may exceed the 60 second limit for Alert Production. A policy is needed on whether Alerts from crowded fields should be allowed a delay, or allowed to be incomplete. K.-T. reports that the control system can easily kill processes that are running over time and move forward with existing outputs -- but that perhaps it will be just as easy to let it keep running and elastically grab additional NCSA resources as needed. This appears possible because the batch system is larger than the Level 1 allocation, although we might not know that this is possible until Level 1 integration happens at NCSA (see Document LDM-230, the operations concepts).

@@

\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{NSV-1} The purpose of this ticket is to ensure that DM's capability for processing non-standard visit images, and required by LSR-REQ-0111, will meet the needs of Special Programs data (which might be considerably more diverse than data for the wide-fast-deep main survey). For now we have adopted a broad interpretation of ``processing" to mean both the initial ISR stage and also the Level 1 Alert Production pipeline, because allowing as many LSST images as possible to contribute to the Alert Stream is a benefit to time domain science. After comparing the anticipated potential diversity in Special Programs data with the currently planned DM pipelines, we have identified several open questions that need to be resolved in advance of the call for Special Programs white papers in the spring of 2018. (1) What is the anticipated minimum (and maximum, if applicable) exposure time for an image to be successfully processed by ISR and DIA? In the latter case, this will probably be limited by the time needed to generate a well-formed PSF. (2) Presently, OSS-REQ-0291 requires that even short exposures be spaced by 15 seconds. Might this requirement be slackened to accommodate a higher cadence Special Program? If so, and if Level 1 processing is required for the science goals, then DM will need to assess options to elastically increase NCSA computational resources and/or the policy for dropping {\it vs.} delaying Alerts during observations with a fast ingest cadence. These issues apply also to Special Programs in the Galactic plane, where crowded fields will increase the volume of Alerts. Alternatively, we could formally require that only 30 second visits will be allowed to contribute to Level 1 Alert Production (3) Long sequences of images on a single field are a likely Special Programs mode (e.g., DDF). We need to confirm that the processing pipeline has the ability to re-parse a scheduled $N$-exposure visit into standard $2$-exposure visits. Consequential images of the same field may lead to unassociated {\tt DIASources} of the same {\tt DIAObject}, and it should be confirmed that the DM pipelines can recognize this when it happens, and flag the resulting Alerts appropriately. \end{enumerate}

@@

{\bf Templates:} The template images that are used in the Level 1 difference imaging pipeline will be built from the Level 2 DRP, and so the first factor affecting a Special Programs image's suitability for Level 1 is to be in a region of sky with an existing template. So long as there is a template, when the exposure time is equivalent to a WFD visit image, $\sim 30$ seconds, treating the image as Level 1 is going to be fine. When processing Special Programs data with the Level 1 pipeline, certain science cases might call for the capability to load and use a certain template that is different from the WFD template (i.e., built over a different timescale). K.-T. confirms that there is not enough memory allotted to store more than one template over the whole sky, but for sub-regions, storing and using an alternative template should be possible. This is not an issue limited to Special Programs, since during commissioning it is conceivable that multiple template versions. K.-T. also confirms that the capability for the processing pipeline to choose a given template based on the programID in the raw image metadata will exist.

@@

The WFD main survey observations will be optimized to meet the mandated goals for moving object discoveries, so it is unclear whether the science goals of MOPS can be much improved by incorporating Special Programs data (aside from e.g., the North Ecliptic Spur area, which has been treated as a Mini-Survey region that will generate a larger number of SSO discoveries). Since MOPS takes {\tt DIASources} as input, any Special Programs images that can be run through the Alert Pipeline can be ingested by MOPS. As discussed under "Solar System Objects (SSO)" in Section \ref{ssec:data_science}, most of the Special Programs data associated with SSO science will obtain standard visit images anyway. There was some concern that a large number of small-separation sources might overwhelm the processing system (i.e., from a deep drilling field with many exposures in a sequence), but upon further consideration this worry was rejected.

@@

Unlike with transient detection and the Level 1 Alert Stream, it is not clear whether it is beneficial to the science goals that depend on the Level 2 DRP to include as many LSST images as possible. We propose that the simplest option is probably that LSST DM deliver a single Level 2 DRP {\tt Objects} database that is built from a set of images with nearly constant depth and cadence. When Special Programs data brings additional area up to the same level of depth and cadence as the rest of the WFD main survey, or improves the Level 2 DRP in some other way, it can be included in the Level 2 DRP. For example, photometric calibrations may require that some or all of the (shallower) Galactic Plane Special Program survey area be incorporated in order to suppress edge effects and low-order modes in the photometric solutions. In short, DM should probably not promise to incorporate any specific data into the Level 2 DRP images or catalogs, and it will not be the default to include Special Programs data into the products for the WFD main survey. This is mostly discussed here in context with the {\tt Source} and {\tt Object} catalogs and the deep CoAdds, but also applies to the DIA pipeline products.

@@

In a ``possible" scenario, DM would assemble a pipeline from existing DM codes in order to process data associated with a given Special Program and build image and catalog products that meet the science needs of that particular program. For example, for a DDF SN survey (see Section \ref{ssec:SPCS_SNDDF}), existing DM codes would be used to: (1) make a deep template image from a certain time window, (2) process standard single visit images, (3) create a nightly CoAdd, (4) run difference imaging analysis, (5) run source detection on the difference images, and (6) create {\tt DIASource} and {\tt DIAObject} catalog equivalents (this example is also given in Section 6 of the DPDD, \cite{LSE-163}). This type of reconfiguration is also be possible for users to do as a Level 3 pipeline (see below), but having these products provided by DM ensures a consistent and verified level of quality and public access to all processed Special Programs data products, and is expected to be a large benefit to the science results.

@@

However, if a Special Program requires processing that cannot be accomplished by reconfiguring existing DM algorithms and codes, then the science community takes responsibility for data processing (i.e., it falls under Level 3). LSST has furthermore allocated an {\it additional} 10\% of its computing resources for user-driven processing and analysis. This 10\% is unrelated to the fraction of Special Programs data, and covers the user-driven reprocessing of any and all LSST data (WFD main survey and Special Programs).

@@

However, in the case where user-driven processing pipelines consume a {\bf large} amount of that 10\%, data export and processing on external computation resources {\bf will} be necessary. The boundaries of what constitutes ``intensive" processing needs are discussed in Section \ref{ssec:docrev_sizing}.

{\bf Level 3 Alert Stream? --} User-driven processing cannot contribute to the Level 1 Alert Stream, but users could create an output that mimics an Alert package. K.-T. raises the issue that running servers or publishing VOEvents from DACs may be tricky due to network and security considerations, but that otherwise publishing in non-LSST Alert-like packets ought to be fine.

{\bf Lower Latency for Data Access? --} There is currently a limit of 24h before processed single visit images will be available, but Special Programs processing might need them sooner. Furthermore, download limits might be imposed and it isn't clear whether it will be possible for a user to pull over the full nightly amount of data if, e.g., they can fund their own server system to process it.

@@

$\bullet$ LSR-REQ-0102 defines the minimum interval between data releases to be {\tt DRT1} = 1 year (1.4.2, Data Release Processing).
\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{LSR-1} RFC (LSE-29): We propose to add a requirement on the schedule or latency for delivering Special Programs data products that is equivalent to the requirements for delivery of yearly data releases (LSR-REQ-0102). E.g., that data products from reconfigured pipelines will be made available on a intermediate time scale where the science goals require it. \end{enumerate}

$\bullet$ LSR-REQ-0111 requires that LSST be capable of obtaining \textit{and processing} exposures that were not taken in a standard visit mode, including those with minimum exposure time of {\tt minExpTime} = 1 second, with the caveat that "non-standard visit exposures may possibly be degraded in some aspects of performance (e.g., cosmic ray rejection)" (2.3.1.2, Non-Standard Visit). The full capabilities of DM to process non-standard visits is not yet well defined; see \ref{NSV-1}.
\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{LSR-2} RFC (LSE-29): We propose that LSR-REQ-0111, the requirement for DM to be capable of processing exposures that were not taken in a standard visit mode, might need to include a caveat once the DM capabilities for processing non-standard visits are truly defined (e.g., extremely short exposures). \end{enumerate}

@@

$\bullet$ LSR-REQ-0055 requires that {\tt userComputingFraction} = 10\% of the total LSST data processing capacity and storage space be allocated for user analysis and Level 3 pipelines and products (2.5.6, Community Computing Services). This is discussed further in Section \ref{ssec:docrev_sizing}.

@@

$\bullet$ OSS-REQ-0027 requires that the scheduling system be able to optimize over at least {\tt nSciProp} = 6 "science proposals", where these "proposals" are observing targets/constraints such as the distribution of filters, the astronomical conditions, and relative priority (OSS 2.1.1.2, Multiple Science Programs).
\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{OSS-1} The purpose of this ticket is to confirm that the minimum number of "science proposals" that can be optimized by the scheduler, {\tt nSciProp} = 6 from OSS-REQ-0027, is not too low, given that some proposed mini-surveys might have quite complicated observing strategies. \end{enumerate} % Assign to Zeljko.

@@

$\bullet$ OSS-REQ-0189 and OSS-REQ-0190 set the minimum number of raw exposures to be supported as {\tt nRawExpNightWinterAvg} = 1960 per night on average (but up to {\tt nRawExpNightMax} = 2800 per night if e.g., two hours of a short-exposure twilight mini-survey are included) and {\tt nRawExpYear} = 5.5$\times10^5$ per year, respectively. These numbers are set by predicting the maximum number of exposures that would be acquired on the longest night of the year in WFD cadence with 2 second slews, assuming $\sim80\%$ completion, but adding a 10\% margin. These estimates appear adequate for Special Programs in general.

$\bullet$ OSS-REQ-0194 and OSS-REQ-0323 set the minimum number of calibration exposures to be supported as {\tt nCalibExpDay} = 450 on average and {\tt nCalExpYear} = 1.5$\times10^5$ per year, respectively. These are {\it minimums}, and so if a Special Program requires additional exposures, this should be possible to accommodate.

>>>>

\setDocRef{DMTN-065}
\date{\today}
\setDocRevision{TBD}
\setDocStatus{draft}

\setDocAbstract{\textbf{This is a draft document and may yet change.} The purpose of this study is to ensure that the LSST Data Management's plans for processing data from Special Programs (Deep Drilling Fields and Mini-Surveys) will meet the needs of the scientific community, and that these plans have been appropriately sized for computational processing and are accurately represented in the work-hours budget. Unresolved issues spawn \textbf{Actions}, which are typically documentation change requests or proposed studies to assess a potential problem.}

@@

\noindent The purpose of this study is to ensure that the requirements and plans of DM are accurately expressed in all major DM documentation, and that the computational system and human work-hours needed to accomplish Special Programs processing have been accurately evaluated. This study spawns \textbf{Action} items, which are typically either documentation change requests or proposals for future studies needed to resolve an issue.

%In particular, this study is designed to review: (1) the processing and products that are required to enable science for Special Programs, (2) the current plans for DM processing and products with respect to Special Programs, (3) whether the DM plans meet the expected needs of Special Programs, and (4) any necessary changes to be written into the requirements, designs, and plans, in which case an \textbf{Action} is spawned.

@@

Section \ref{sec:docrev} steps through each of the relevant LSST documents and highlights specific requirements or text in these documents that might need additions or clarifications to properly reflect DM's plans for processing Special Programs data.
%These \textbf{Concerns} are not prioritized by urgency or importance, but instead represent more of a complete set of possible issues. \textbf{The Concerns raised by this study fall into two main categories:} \\
%1. Whether DM will be able to process the full diversity of non-standard visit images, and if not, what the boundaries will be. \\
%2. Ensuring that DM's intent to process Special Programs data with reconfigured pipelines, and to provide separate data products that meet the science goals of each program, is accurately reflected in all documents, as it is in the DPDD, and is in the staffing budget. \\

@@

All aspects related to DM's \textit{processing} of Special Programs data are addressed in Section \ref{sec:dmplans}.

@@

\textbf{Shorter Exposures.} The camera constraint on the minimum supported exposure time is currently 1 second (stretch goal 0.1 seconds). The minimum exposure time for an image to be successfully reduced with Instrument Signature Removal (ISR) will be a DM constraint. Assuming that 1 second exposure can be reduced and calibrated, its detected point sources will span a dynamic range of $r \approx 12.9$ -- $21.0$ magnitudes. A template image built on $15$ second exposures will saturate at $r \approx 15.8$, but this still leaves stars between $15.8$--$21.0$ magnitudes to be used in the PSF-matching (and all other filters have a similarly large overlap). However, in order for an image to be successfully PSF-matched to the template, the PSF must be well formed (i.e., no speckle pattern), and have a spatial variance that the pipeline is capable of modeling (i.e., be smoothly varying on some minimal scale). As a simple demonstration, MLG has made Figure \ref{fig:expt} to show that perhaps exposure times shorter than 2 seconds do not have a well-formed PSF (using the centroid of a 2D Gaussian fit as a proxy for "well-formed"). Another concern is that the Level 1 pipeline would not be able to deliver the required $60$ second latency if the camera produces more than 1 image in 30 seconds, which would disqualify a contribution to the Alert Stream.

@@

\textbf{Longer Exposures.} There is no maximum exposure time specified for an LSST image. Given that the template image will be a stack of at least a year or two of data, processing a $5$--$10$ times deeper single image through the difference imaging pipeline should be fine. However, a $2\times150$ second exposure would saturate at $r \approx 18.3$, and cosmic-ray rejection completeness might suffer (unknown), which could impact the quality of a difference image and the detected sources. With a longer exposure time the $60$ second latency requirement could still be met, but any system qualities that vary on short (but $>30$ second) timescales could inhibit photometric calibration (e.g., tracking).

In conversation with DM-AP team members (Reiss, Findeisen, Connolly, Bo) there has not yet been a study of the safe range of exposure times that will be allowed to contribute to the Level 1 and Alert Stream. One possibly useful study is Chang et al. (2012), "Atmospheric point spread function interpolation for weak lensing in short exposure imaging data". They show that a 15 second exposure contains PSF variability on short spatial scales across a 1 square degree image which, for extragalactic fields with few stars (i.e., but good for weak lensing), is hard to characterize. They present a new software package to do mitigate the effects. Alternatively, we may need to use software packages \texttt{PhoSim} (Peterson et al. 2015; \url{http://adsabs.harvard.edu/abs/2015ApJS..218...14P}) or \texttt{ARROYO} (\url{http://proceedings.spiedigitallibrary.org/proceeding.aspx?articleid=848256}) to at least simply characterize the PSF stability as a function of exposure time.

@@

OSS-REQ-0291 requires that short exposures be spaced by 15 seconds, and so DM's ingest rate of exposures should never be more than 1 per 15 seconds, or a visit rate of 1 per 30 seconds. However, \textit{if a higher image acquisition rate is cleared by the camera team}, then DM might encounter some issues such as the minimum time required to save an image to disk, the handling of a processing backup, and/or inconsistently associated \texttt{DIASources}.

\textbf{Minimum Save Time:} If short exposures are taken in an unbuffered sequence, the minimum time between exposures is the shutter-crossing time (1 second) plus the readout time (2 seconds): 3 seconds. This is too rapid an acquisition rate for processing with Level 1, but what if the science goals accept saving them to disk and analyzing them later? \textbf{How long does it take to save an image to disk?}

\textbf{Processing backup:} Are there options for short-term increases in parallel processing power at NCSA? (Such options might be needed anyway to process crowded fields with $>$10k \texttt{DIASources}.) The bandwidth needed to load templates at a faster rate is also a concern, since templates have twice as many pixels, if the data acquistion rate is surveying at twice the WFD main survey rate, that's a $4\times$ additional bandwidth load. However, at NCSA there should be ways to elastically change the amount of processing power available. This is also an issue for crowded fields (below).

\textbf{Unassociated \texttt{DIASources}:} the consistency of \texttt{DIAObjects} suffers if the same patch of sky is observed in quick succession (overlapping regions of short exposures and/or sustained 30 second visits to the same field). In other words, it is not guaranteed that the next string of processing would have access to the latest \texttt{DIAObjects} for association. It is furthermore unclear whether we can flag an Alert as potentially having incomplete source-object associations. This is also an issue for crowded fields (below).

@@

K.-T. Lim has pointed out that an odd number of exposures is a non-standard visit; two snaps is hardwired into the code. This is baked-in to a configuration so that the pipeline can have a definition of what kind of timing delay constitutes ``late".  Moving away from 2 exposures per visit requires a configuration change to the pipelines, which incurs an overhead (up to 1 minute) -- in fact, K.-T. things that between $10$ and $120$ seconds exposure times can easily be handled by the pipeline (i.e., can be run through ISR using scaled calibration frames), so long as they come in pairs. The real problem is knowing how long the processing should take, and not killing a process that is taking longer because there were 4 snaps in the visit instead of 2. To accommodate non-standard visits requires that the scheduler pass on the information of the number of snaps in the visit (\ref{DMSR-1}). Then the processing pipeline will know to, e.g., not attempt to difference the two snaps in the case were there is an odd number of snaps in a visit. \textit{MLG -- I've heard rumors of a CR regarding alternate standard visits of $1\times30$ seconds, but do not know the status or implications of this.}

@@

Special Programs are more likely to include crowded fields than the WFD main survey area. Due to the increased number of sources, the number of \texttt{DIASources} -- and therefore the number of Alerts -- increases as well. In turn, this increases the processing time and in some cases, may exceed the 60 second limit for Alert Production. A policy is needed on whether Alerts from crowded fields should be allowed a delay, or allowed to be incomplete. K.-T. reports that the control system can easily kill processes that are running over time and move forward with existing outputs -- but that perhaps it will be just as easy to let it keep running and elastically grab additional NCSA resources as needed. This appears possible because the batch system is larger than the Level 1 allocation, although we might not know that this is possible until Level 1 integration happens at NCSA (see Document LDM-230, the operations concepts).

@@

\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{NSV-1} The purpose of this ticket is to ensure that DM's capability for processing non-standard visit images, and required by LSR-REQ-0111, will meet the needs of Special Programs data (which might be considerably more diverse than data for the wide-fast-deep main survey). For now we have adopted a broad interpretation of ``processing" to mean both the initial ISR stage and also the Level 1 Alert Production pipeline, because allowing as many LSST images as possible to contribute to the Alert Stream is a benefit to time domain science. After comparing the anticipated potential diversity in Special Programs data with the currently planned DM pipelines, we have identified several open questions that need to be resolved in advance of the call for Special Programs white papers in the spring of 2018. (1) What is the anticipated minimum (and maximum, if applicable) exposure time for an image to be successfully processed by ISR and DIA? In the latter case, this will probably be limited by the time needed to generate a well-formed PSF. (2) Presently, OSS-REQ-0291 requires that even short exposures be spaced by 15 seconds. Might this requirement be slackened to accommodate a higher cadence Special Program? If so, and if Level 1 processing is required for the science goals, then DM will need to assess options to elastically increase NCSA computational resources and/or the policy for dropping \textit{vs.} delaying Alerts during observations with a fast ingest cadence. These issues apply also to Special Programs in the Galactic plane, where crowded fields will increase the volume of Alerts. Alternatively, we could formally require that only 30 second visits will be allowed to contribute to Level 1 Alert Production (3) Long sequences of images on a single field are a likely Special Programs mode (e.g., DDF). We need to confirm that the processing pipeline has the ability to re-parse a scheduled $N$-exposure visit into standard $2$-exposure visits. Consequential images of the same field may lead to unassociated \texttt{DIASources} of the same \texttt{DIAObject}, and it should be confirmed that the DM pipelines can recognize this when it happens, and flag the resulting Alerts appropriately. \end{enumerate}

@@

\textbf{Templates:} The template images that are used in the Level 1 difference imaging pipeline will be built from the Level 2 DRP, and so the first factor affecting a Special Programs image's suitability for Level 1 is to be in a region of sky with an existing template. So long as there is a template, when the exposure time is equivalent to a WFD visit image, $\sim 30$ seconds, treating the image as Level 1 is going to be fine. When processing Special Programs data with the Level 1 pipeline, certain science cases might call for the capability to load and use a certain template that is different from the WFD template (i.e., built over a different timescale). K.-T. confirms that there is not enough memory allotted to store more than one template over the whole sky, but for sub-regions, storing and using an alternative template should be possible. This is not an issue limited to Special Programs, since during commissioning it is conceivable that multiple template versions. K.-T. also confirms that the capability for the processing pipeline to choose a given template based on the programID in the raw image metadata will exist.

@@

The WFD main survey observations will be optimized to meet the mandated goals for moving object discoveries, so it is unclear whether the science goals of MOPS can be much improved by incorporating Special Programs data (aside from e.g., the North Ecliptic Spur area, which has been treated as a Mini-Survey region that will generate a larger number of SSO discoveries). Since MOPS takes \texttt{DIASources} as input, any Special Programs images that can be run through the Alert Pipeline can be ingested by MOPS. As discussed under "Solar System Objects (SSO)" in Section \ref{ssec:data_science}, most of the Special Programs data associated with SSO science will obtain standard visit images anyway. There was some concern that a large number of small-separation sources might overwhelm the processing system (i.e., from a deep drilling field with many exposures in a sequence), but upon further consideration this worry was rejected.

@@

Unlike with transient detection and the Level 1 Alert Stream, it is not clear whether it is beneficial to the science goals that depend on the Level 2 DRP to include as many LSST images as possible. We propose that the simplest option is probably that LSST DM deliver a single Level 2 DRP \texttt{Objects} database that is built from a set of images with nearly constant depth and cadence. When Special Programs data brings additional area up to the same level of depth and cadence as the rest of the WFD main survey, or improves the Level 2 DRP in some other way, it can be included in the Level 2 DRP. For example, photometric calibrations may require that some or all of the (shallower) Galactic Plane Special Program survey area be incorporated in order to suppress edge effects and low-order modes in the photometric solutions. In short, DM should probably not promise to incorporate any specific data into the Level 2 DRP images or catalogs, and it will not be the default to include Special Programs data into the products for the WFD main survey. This is mostly discussed here in context with the \texttt{Source} and \texttt{Object} catalogs and the deep CoAdds, but also applies to the DIA pipeline products.

@@

In a ``possible" scenario, DM would assemble a pipeline from existing DM codes in order to process data associated with a given Special Program and build image and catalog products that meet the science needs of that particular program. For example, for a DDF SN survey (see Section \ref{ssec:SPCS_SNDDF}), existing DM codes would be used to: (1) make a deep template image from a certain time window, (2) process standard single visit images, (3) create a nightly CoAdd, (4) run difference imaging analysis, (5) run source detection on the difference images, and (6) create \texttt{DIASource} and \texttt{DIAObject} catalog equivalents (this example is also given in Section 6 of the DPDD, \cite{LSE-163}). This type of reconfiguration is also be possible for users to do as a Level 3 pipeline (see below), but having these products provided by DM ensures a consistent and verified level of quality and public access to all processed Special Programs data products, and is expected to be a large benefit to the science results.

@@

However, if a Special Program requires processing that cannot be accomplished by reconfiguring existing DM algorithms and codes, then the science community takes responsibility for data processing (i.e., it falls under Level 3). LSST has furthermore allocated an \textit{additional} 10\% of its computing resources for user-driven processing and analysis. This 10\% is unrelated to the fraction of Special Programs data, and covers the user-driven reprocessing of any and all LSST data (WFD main survey and Special Programs).

@@

However, in the case where user-driven processing pipelines consume a \textbf{large} amount of that 10\%, data export and processing on external computation resources \textbf{will} be necessary. The boundaries of what constitutes ``intensive" processing needs are discussed in Section \ref{ssec:docrev_sizing}.

\textbf{Level 3 Alert Stream? --} User-driven processing cannot contribute to the Level 1 Alert Stream, but users could create an output that mimics an Alert package. K.-T. raises the issue that running servers or publishing VOEvents from DACs may be tricky due to network and security considerations, but that otherwise publishing in non-LSST Alert-like packets ought to be fine.

\textbf{Lower Latency for Data Access? --} There is currently a limit of 24h before processed single visit images will be available, but Special Programs processing might need them sooner. Furthermore, download limits might be imposed and it isn't clear whether it will be possible for a user to pull over the full nightly amount of data if, e.g., they can fund their own server system to process it.

@@

$\bullet$ LSR-REQ-0102 defines the minimum interval between data releases to be \texttt{DRT1} = 1 year (1.4.2, Data Release Processing).
\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{LSR-1} RFC (LSE-29): We propose to add a requirement on the schedule or latency for delivering Special Programs data products that is equivalent to the requirements for delivery of yearly data releases (LSR-REQ-0102). E.g., that data products from reconfigured pipelines will be made available on a intermediate time scale where the science goals require it. \end{enumerate}

$\bullet$ LSR-REQ-0111 requires that LSST be capable of obtaining \textit{and processing} exposures that were not taken in a standard visit mode, including those with minimum exposure time of \texttt{minExpTime} = 1 second, with the caveat that "non-standard visit exposures may possibly be degraded in some aspects of performance (e.g., cosmic ray rejection)" (2.3.1.2, Non-Standard Visit). The full capabilities of DM to process non-standard visits is not yet well defined; see \ref{NSV-1}.
\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{LSR-2} RFC (LSE-29): We propose that LSR-REQ-0111, the requirement for DM to be capable of processing exposures that were not taken in a standard visit mode, might need to include a caveat once the DM capabilities for processing non-standard visits are truly defined (e.g., extremely short exposures). \end{enumerate}

@@

$\bullet$ LSR-REQ-0055 requires that \texttt{userComputingFraction} = 10\% of the total LSST data processing capacity and storage space be allocated for user analysis and Level 3 pipelines and products (2.5.6, Community Computing Services). This is discussed further in Section \ref{ssec:docrev_sizing}.

@@

$\bullet$ OSS-REQ-0027 requires that the scheduling system be able to optimize over at least \texttt{nSciProp} = 6 "science proposals", where these "proposals" are observing targets/constraints such as the distribution of filters, the astronomical conditions, and relative priority (OSS 2.1.1.2, Multiple Science Programs).
\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{OSS-1} The purpose of this ticket is to confirm that the minimum number of "science proposals" that can be optimized by the scheduler, \texttt{nSciProp} = 6 from OSS-REQ-0027, is not too low, given that some proposed mini-surveys might have quite complicated observing strategies. \end{enumerate} % Assign to Zeljko.

@@

$\bullet$ OSS-REQ-0189 and OSS-REQ-0190 set the minimum number of raw exposures to be supported as \texttt{nRawExpNightWinterAvg} = 1960 per night on average (but up to \texttt{nRawExpNightMax} = 2800 per night if e.g., two hours of a short-exposure twilight mini-survey are included) and \texttt{nRawExpYear} = 5.5$\times10^5$ per year, respectively. These numbers are set by predicting the maximum number of exposures that would be acquired on the longest night of the year in WFD cadence with 2 second slews, assuming $\sim80\%$ completion, but adding a 10\% margin. These estimates appear adequate for Special Programs in general.

$\bullet$ OSS-REQ-0194 and OSS-REQ-0323 set the minimum number of calibration exposures to be supported as \texttt{nCalibExpDay} = 450 on average and \texttt{nCalExpYear} = 1.5$\times10^5$ per year, respectively. These are \textit{minimums}, and so if a Special Program requires additional exposures, this should be possible to accommodate.

$\bullet$ Section 3.1.5 (OSS-REQ-0125 and those following) set the requirements for data products, including their variety and format (i.e., images and catalogs in Level 1 and 2), their precision (e.g., photometric and astrometric accuracy, completeness/spurious thresholds for DIA sources), and the timeline for the release of data products (Alerts, Level 1, and Level 2 ). Would these also apply to data from Special Programs? For example, OSS-REQ-0157 sets the fraction of false detections in deep CoAdds caused by unremoved artifacts to be \texttt{falseDeepDetect} = 0.1\%, but would this apply to a DDF stack?
\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{OSS-3} RFC (LSE-30): We propose to add a statement to Section 3.1.5 of the OSS that DM will make a ``best effort" to meet the data products requirements with the Special Programs data. \end{enumerate}
